A Review of Mortality Models

### 2020-09-07 mortality-models/draft.tex
Electrodynamics

\section{Makeham's law}\label{s3}
William Makeham, the English mathematician and actuary, added a term to
Gompertz's law of resistance to death which the latter had acknowledged but
not included. Makeham's modification \cite{makeham1867law} of equation 
\eqref{s2e3} is
\begin{equation}\label{s3e1}
\mu(x) = A + Be^{kx},
\end{equation}
where $A$ is a constant independent of the person's age. Makeham's survival
function is
\begin{equation}\label{s3e2}
s(x) = C\exp\left(Ax + \frac{B}{k}\exp(kx)\right).
\end{equation}
Makeham's survival function has three constants $A, B$ and $C$ and therefore
can be fitted to emprically observed mortality data more effectively than
Gompertz's model. Makeham's law was often applied to mortality tables in the
age range beyond $20$ years to almost end of life\cite{jordan1967society}.
\bibliography{draft}
\bibliographystyle{plain}
\end{document}

>>>>

\section{Makeham's law and its generalisations}\label{s3}
William Makeham, the English mathematician and actuary, added a term to
Gompertz's law of resistance to death which the latter had acknowledged but
not included. Makeham's modification \cite{makeham1867law} of equation 
\eqref{s2e3} is
\begin{equation}\label{s3e1}
\mu(x) = A + Be^{kx},
\end{equation}
where $A$ is a constant independent of the person's age. Makeham's survival
function is
\begin{equation}\label{s3e2}
s(x) = C\exp\left(Ax + \frac{B}{k}\exp(kx)\right).
\end{equation}
Makeham's survival function has three constants $A, B$ and $C$ and therefore
can be fitted to emprically observed mortality data more effectively than
Gompertz's model. Makeham's law was often applied to mortality tables in the
age range beyond $20$ years to almost end of life\cite{jordan1967society}. 
In 1890\cite{makeham1890further}, Makeham proposed yet another modification to 
his law by adding a linear term to the hazard function or the force of 
mortality. Makeham thought that in addition to the two factors spotted by
Gompertz there is another one that grows linearly with age. It was proposed to 
be
\begin{equation}\label{s3e3}
\mu(x) = A_1 + A_2x + Be^{kx}.
\end{equation}
The resulting survival function is
\begin{equation}\label{s3e4}
s(x) = C\exp\left(A_1x + \frac{A_2}{2}x^2 + \frac{B}{k}\exp(kx)\right).
\end{equation}
Makeham's law, like its predecessor, Gompertz's law was used to construct
mortality tables of adults. Their simplicity allowed them to be used 
effectively up to as late as the 1940s. Gompertz law was used to construct
the 1937 Standard Annuity Table and Makeham's law for 1940 Annuity table
\cite{jordan1967society}. Both laws assumed that the force of mortality 
increases monotonically with time. However, it is commonly observed that 
infant mortality contradicts this assumption. In order to extend Makeham's
law of 1860 to infantile ages, Wilhem Lazarus \cite{lazarus1867ueber} added
an exponentially decreasing term. His law expressed the force of mortality
as 
\begin{equation}\label{s3e5}
\mu(x) = B_1 e^{-k_1x} + A + B_2e^{k_2 x}.
\end{equation}
The constants $B_1$ and $B_2$ are positive numbers. The corresponding survival 
function is
\begin{equation}\label{s3e6}
s(x) = 
C\exp\left(-\frac{B_1}{k_1}e^{-k_1x} + Ax + \frac{B_2}{k_2}e^{k_2x}\right)
\end{equation}

The way infants' susceptibility to die of diseases upsets Gompertz's and 
Makeham's assumption of monotonicity of the law of mortality, do does the 
spike in the death rate of young adults. Wars were frequent in Europe of the 
eighteenth century and the victims were disproportionately the young recruits.
Even otherwise, young adults are more prone to death by accidents. In order
to take into account this mortality hump in the young, Thiele 
\cite{thiele1871mathematical} introduced a gaussian term to Lazarus' model.
His law of mortality is
\begin{equation}\label{s3e7}
\mu(x) = B_1 e^{-k_1x} + Ae^{-a_1(x - a_2)^2} + B_2e^{k_2 x}.
\end{equation}
All the constants in this equation, namely $A, a_1, a_2, B_1, k_1, B_2$ and 
$k_2$ are positive numbers. The presence of the gaussian term prevents us
from writing a closed form expression of the survival function. However, the
ordinary differential equation for the survival function can be readily 
integrated numerically. Alternatively, if the initial conditions are available,
one can write its solution in terms of the error function.

We observe that all generalisations of Makeham's laws introduce additional
constants to the force of mortality and the hazard function. These constants
provide additional `degrees of freedom' to the functions used to fit the 
mortality data, resulting in a better fit.

Forfar, McCutcheon and Wilkie proposed a far wider generalisation of the
Gompertz and Makeham's laws. They defined a class of models, called GM models
after their inventors, with the mathematical form
\begin{equation}\label{s3e8}
\mu(x) = \sum_{i=0}^{r-1}\alpha_i x^i + 
\exp\left(\sum_{i=1}^{s-1}\beta_i x^i\right).
\end{equation}
They also used the convention that if $r = 0$ then the polynomial terms are 
absent and if $s = 0$ then the exponential terms do not appear. They called
the law of force of mortality of equation \eqref{s3e8} as $\mathrm{GM}(r,s)$.
It is easy to check that Gompertz's law is $\mathrm{GM}(0, 2)$, Makeham's
first law is $\mathrm{GM}(1,2)$ and Makeham's second law is $\mathrm{GM}(2,2)$.
The $\mathrm{GM}(0, 2)$, $\mathrm{GM}(2,2)$ and $\mathrm{GM}(1,3)$ are used
by the Continuous Mortality Investigation Bureau\cite{pitacco2016high} in the 
United Kingdom.

\section{Other mortality laws prior to the 20th century}\label{s4}
de Moivre's, Gompertz's and Makeham's laws of mortality were not the only
ones in vogue in the nineteenth century and prior to it. In this section we
mention a few notable ones without commenting on their merits. Lambert's
survival function \cite{lambert1776dottrina} was
\begin{equation}\label{s4e1}
s(x) = \left(\frac{a - x}{x}\right)^2 - b\left(e^{-x/c} - e^{-x/d}\right).
\end{equation}
It had four degrees of freedom. Babbage\cite{babbage1823tables} proposed a 
quadratic survival function
\begin{equation}\label{s4e2}
s(x) = c - bx - ax^2.
\end{equation}
Young proposed a polynomial of degree $40$ as the survival function in 
\cite{adler1866memoir} 1826. von Littrow\cite{von1832lebensversicherungen}
extended Babbage's model to polynomials of higher degree. Moser
\cite{moser1839gesetze} proposed a polynomial survival function with 
fractional powers of $x$. It has five degrees of freedom and had the form
\begin{equation}\label{s4e3}
s(x) = 1 - ax^{1/4} _ bx^{9/4} - cx^{17/4} - dx^{25/4} + ex^{33/4}.
\end{equation}
All these models were attempts to fit the mortality data with a continuous 
function. They were not rooted in demographic observations like the models
of Gompertz, Makeham, Lazarus or Thiele. As a result, they soon fell into
disuse after Gompertz proposed his model and others improved upon it.
\bibliography{draft}
\bibliographystyle{plain}
\end{document}
